---
title: "Data Visualization: Day 3"
author: "Erik Westlund"
date: "2025-06-12"
date-modified: "`r format(Sys.Date(), '%Y-%m-%d')`"
---

```{r setup}
#| echo: false
#| message: false
#| warning: false

library(dplyr)
library(purrr)
library(janitor)
library(dagitty)
library(forcats)
library(ggdag)
library(ggplot2)
library(ggtext)
library(stringr)
library(kableExtra)
```


# Day 3: What to Visualize – From Question to Plot

## Overview for Day 3

::: {.incremental}
- Scientific questions and DAGs
- What to visualize: distributions, time, space, relationships, models
- Key design principles: honesty, annotation, small multiples
- Case studies: 7 applied examples
:::


## Theme: Science Drives Visualization

* Data visualization is valuable for its scientific application
* It helps us:
    - Understand our data
    - Understand our models
    - Communicate our findings in an impactful way

# DAGs, Scientific Complexity, Causal Inference, and Visualization

## From Question to DAG to Plot

* We'll start with a scientific question.
* We'll sketch out the causal structure using a DAG.
* From this, we can assess typical problems scientific studies like this face
* We can then use the DAG to guide our visualization choices

## Scientific Question

"What causes mothers to receive comprehensive postnatal care?"

## Directed Acyclic Graphs

* A DAG is a graph that represents the causal relationships between variables.
* It is a directed graph, meaning that the edges have a direction.
* It is an acyclic graph, meaning that there are no cycles in the graph.
* It is a graph, meaning that it has nodes and edges.

## How DAGs Help Us In Science &amp; Visualization

* DAGs help us:

    - Identify the variables that are important to the question
    - Identify the variables that are confounding
    - Identify the variables that are missing
    - Identify the variables that are measured incorrectly

* Having a good idea of the causal structure of a question helps us know what to visualize.

## DAGs as a Visualization Tool

* DAGs:

    1. are themselves visualizations
    2. encode scientific problems and causal structures
* We can use the `daggity` R package to visualize the the causal structure of the problems around the motivating research question

## Big Picture

:::: {.columns}
::: {.column width="60%"}
```{r causes_general}
rcp_dag <- dagify(RCP ~ PQ + PC + WTP + PT + RA + RP)

rcp_dag |>
    ggdag() +
    theme_dag()
```
:::

::: {.column width="40%" .small}
- RCP = "Received Comprehensive Postnatal Care"
- PQ = "Provider Quality"
- PC = "Personal Capacity"
- WTP = "Willingness to Pay"
- PT = "Provider Trust"
- RA = "Risk Aversion"
- RP = "Risk Profile"
:::
::::

## Provider Quality (PQ) Sub-DAG

:::: {.columns}
::: {.column width="60%"}
```{r provider_quality}
pq_dag <- dagify(
  PQ ~ PEC,
  PEC ~ ST
)

pq_dag |>
    ggdag() +
    theme_dag()
```
:::

::: {.column width="40%" .small}
- PQ = "Provider Quality"
- PEC = "Political/Economic Conditions"
- ST = "State"
:::
::::

## Personal Capacity (PC) Sub-DAG

:::: {.columns}
::: {.column width="60%"}
```{r personal_capacity}
pc_dag <- dagify(
  PC ~ DEP + JOB + INC + DIS,
  DEP ~ INC + JOB + CO,
  JOB ~ EDU + CO + R + M + I + CC,
  DIS ~ PEC,
  INC ~ JOB + EDU + RE + PEC + AGE
)

pc_dag |>
    ggdag() +
    theme_dag()
```
:::

::: {.column width="40%" .small}
- PC = "Personal Capacity"
- DEP = "Dependents"
- JOB = "Job Type"
- INC = "Income"
- DIS = "Distance to Provider"
- CO = "Cultural Orientation"
- EDU = "Educational Attainment"
- R = "Resilience"
- M = "Motivation"
- I = "Intelligence"
- CC = "Community Connections"
- RE = "Race/Ethnicity"
- AGE = "Age"
:::
::::

## Willingness to Pay (WTP) Sub-DAG

:::: {.columns}
::: {.column width="60%"}
```{r willingness_to_pay}
wtp_dag <- dagify(
  WTP ~ PQ + INC + INS + RA + CO,
  INS ~ JOB + PEC + AGE,
  CO ~ PED + PIN + REL + PEC + CC
)

wtp_dag |>
    ggdag() +
    theme_dag()
```
:::

::: {.column width="40%" .small}
- WTP = "Willingness to Pay"
- PQ = "Provider Quality"
- INC = "Income"
- INS = "Insurance"
- RA = "Risk Aversion"
- CO = "Cultural Orientation"
- JOB = "Job Type"
- PEC = "Political/Economic Conditions"
- AGE = "Age"
- PED = "Parent Education"
- PIN = "Parent Income"
- REL = "Religion"
- CC = "Community Connections"
:::
::::

## Provider Trust (PT) Sub-DAG

:::: {.columns}
::: {.column width="60%"}
```{r provider_trust}
pt_dag <- dagify(
  PT ~ PQ + RE + CO,
  CO ~ PED + PIN + REL + PEC + CC
)

pt_dag |>
    ggdag() +
    theme_dag()
```
:::

::: {.column width="40%" .small}
- PT = "Provider Trust"
- PQ = "Provider Quality"
- RE = "Race/Ethnicity"
- CO = "Cultural Orientation"
- PED = "Parent Education"
- PIN = "Parent Income"
- REL = "Religion"
- PEC = "Political/Economic Conditions"
- CC = "Community Connections"
:::
::::

## Risk Aversion (RA) Sub-DAG

:::: {.columns}
::: {.column width="60%"}
```{r risk_aversion}
ra_dag <- dagify(
  RA ~ PQ + RP + INS,
  RP ~ PQ + AGE + OBE + MG + DM + HD + PP + PR + HT + GHT,
  INS ~ JOB + PEC + AGE
)

ra_dag |>
    ggdag() +
    theme_dag()
```
:::

::: {.column width="40%" .small}
- RA = "Risk Aversion"
- PQ = "Provider Quality"
- RP = "Risk Profile"
- INS = "Insurance"
- AGE = "Age"
- OBE = "Obesity"
- MG = "Multiple Gestation"
- DM = "Diabetes Mellitus"
- HD = "Heart Disease"
- PP = "Placenta Previa"
- PR = "Preeclampsia"
- HT = "Hypertension"
- GHT = "Gestational Hypertension"
- JOB = "Job Type"
- PEC = "Political/Economic Conditions"
:::
::::

## Risk Profile (RP) Sub-DAG

:::: {.columns}
::: {.column width="60%"}
```{r risk_profile}
rp_dag <- dagify(
  RP ~ PQ + AGE + OBE + MG + DM + HD + PP + PR + HT + GHT,
  OBE ~ PEC + AGE,
  MG ~ AGE + OBE,
  DM ~ AGE + OBE + INC,
  HD ~ AGE + OBE + DM,
  PP ~ AGE + MG,
  PR ~ AGE + HT + GHT + MG,
  HT ~ AGE + OBE,
  GHT ~ HT + MG
)

rp_dag |>
    ggdag() +
    theme_dag()
```
:::

::: {.column width="40%" .small}
- RP = "Risk Profile"
- PQ = "Provider Quality"
- AGE = "Age"
- OBE = "Obesity"
- MG = "Multiple Gestation"
- DM = "Diabetes Mellitus"
- HD = "Heart Disease"
- PP = "Placenta Previa"
- PR = "Preeclampsia"
- HT = "Hypertension"
- GHT = "Gestational Hypertension"
- PEC = "Political/Economic Conditions"
- INC = "Income"
:::
::::

## Complete DAG Codification

```{r complete_dag}
complete_dag <- dagify(
  RCP ~ PQ + PC + WTP + PT + RA + RP,
  PQ ~ PEC,
  PC ~ DEP + JOB + INC + DIS,
  WTP ~ PQ + INC + INS + RA + CO,
  PT ~ PQ + RE + CO,
  RA ~ PQ + RP + INS,
  RP ~ PQ + AGE + OBE + MG + DM + HD + PP + PR + HT + GHT,
  PEC ~ ST,
  DEP ~ INC + JOB + CO,
  JOB ~ EDU + CO + R + M + I + CC,
  DIS ~ PEC,
  INC ~ JOB + EDU + RE + PEC + AGE,
  INS ~ JOB + PEC + AGE,
  CO ~ PED + PIN + REL + PEC + CC,
  PIN ~ PEC + PED,
  EDU ~ PED + PIN + R + M + I + CC,
  R ~ PR,
  M ~ PM,
  I ~ PI,
  CC ~ PCC,
  REL ~ RE,
  OBE ~ PEC + AGE,
  MG ~ AGE + OBE,
  DM ~ AGE + OBE + INC,
  HD ~ AGE + OBE + DM,
  PP ~ AGE + MG,
  PR ~ AGE + HT + GHT + MG,
  HT ~ AGE + OBE,
  GHT ~ HT + MG
)

```

## Complete DAG Visualization

```{r complete_dag_visualization}
complete_dag |>
    ggdag() +
    theme_dag()
```

# DAGs and Scientific Complexity

## So What?

* I think it's worth stepping back and considering how complex many scientific questions really are
* We do not always observe everythign we want to, but with the help of DAGs we can:
    - Identify the variables that are important to the question
    - Identify what we can actually observe
    - Identify what we cannot observe
    - Assess the limitations we face with causal analysis

## Next Steps

* We are now going to work through seven applications/case studies, each addressing a different aspect of visualization.
* In our first case study, we will examine how to visualize spatial variation.

## Side Quest: Simulated Data

* In `examples/dag_sim_data.qmd` we simulated data that encodes the DAG structure.
* Let's take a very brief glance at that file to see how we simulated the data.

# Applications

## Application 1: Choropleths

* In `applications_1_choropleth.qmd` we create a choropleth map of the United States using the simulated data from above
* We show how even with maps, small multiples help find clarity

## Application 2: Forest Plots

* In `applications_2_forest_plots.qmd` we create a forest plot of the data using the same simulated data.
* We show how spatial variation can often be visualized better without actual maps.

## Application 3: Ridgeline Plots

* In `applications_3_ridge_plots.qmd` we create a ridgeline plot of the data using the same simulated data.
* We show how ridgeline plots can show distributional differences across groups.

## Application 4: Distributions

* In `applications_4_distributions.qmd` we examine distributions of variables
* Wecreate box plots and violin plots to show the distribution of a variable across a population.

## Application 5: Time Trends

* In `applications_5_time_trends.qmd` we examine time trends of variables
* We create line plots to show how a variable changes over time.

## Application 6: Scale Honesty & Effectiveness  

* In `applications_6_scale_honesty.qmd` we examine how axes can be used to mislead or clarify
* We briefly discuss how what counts as dishonest may be contextual

## Application 7: Visualizing Models  

* In `applications_7_visualizing_models.qmd` we visualize model outputs
* We create coefficient plots and model predictions to help interpret the model


# Summary and Takeaways

* Visualization choices should be driven by the **scientific question**
* Experimentation is key; often the "obvious" plot is not the best plot
* Iterate: build in layers; get the basics down; then polish
* Use small multiples, annotations, and thoughtful scales to guide the viewer