---
title: "AI & LLM Example"
author: "Erik Westlund"
date: "2025-06-11"
date-modified: "`r format(Sys.Date(), '%Y-%m-%d')`"
editor:
  render-on-save: true
---

## Overview

* We'll use an LLM to help us prepare data for a visualization
* Let's use the [CDC's data on influenza vaccination coverage for all ages](https://data.cdc.gov/Flu-Vaccinations/Influenza-Vaccination-Coverage-for-All-Ages-6-Mont/vh55-3he6/about_data)

## Data Preparation

```{r setup}
#| echo: false
#| message: false

library(dplyr)
library(ggplot2)
library(forcats)
library(janitor)
library(kableExtra)
library(readr)
library(stringr)
library(tidyr)

```


```{r}
#| message: false

df <- read_csv(here::here("data", "raw", "cdc_Influenza_Vaccination_Coverage_for_All_Ages__6__Months__20250610.csv"))

df |> glimpse()
```

## Chat with the LLM

### Data Preparation

From here, we can chat with the LLM to help us prepare the data for a visualization.

[Below I provide key aspects of the prompt's sent to the ChatGPT LLM as well as the results. I've excluded some of the LLM's missteps.]



```{markdown}
I am working with an unfamiliar data set

Here is a glimpse using dplyr

what do you see

> df <- read_csv(here::here("data", "raw", "cdc_Influenza_Vaccination_Coverage_for_All_Ages__6__Months__20250610.csv"))
Rows: 220729 Columns: 11                                                                                                                                                                               
── Column specification ───────────────────────────────────────────────────────────────────────────────────────────────────────────────────────────────────────────────────────────────────────────────
Delimiter: ","
chr (9): Vaccine, Geography Type, Geography, FIPS, Season/Survey Year, Dimension Type, Dimension, Estimate (%), 95% CI (%)
dbl (2): Month, Sample Size

ℹ Use spec() to retrieve the full column specification for this data.
ℹ Specify the column types or set show_col_types = FALSE to quiet this message.
> 
> df |> glimpse()
```

The bot then summarized the data set.

Using the following prompt, We then sent the LLM the actual data file for more context (🚨 do not do this with private data! 🚨)

```{markdown}
I am actually going to give you the file.  Interpret it. We are going to use dplyr and ggplot2 to prepare it and visualize aspects of it.

tell me what you see
```

It responded with a summary, warning us about some of the odd structures of the data.

We then asked it to clean the data:

```{markdown}
Yes, let's get clean, standardized column names. Let's make all factor variables factor types. Lets extract the confidence intervals into separate columns.  Let's just start there. Give me R code that I can paste into RStudio.
```

Please note from here on out, ChatGPT gave us code that includes library calls and new code. The library calls are loaded above.

It first gave us back broken code because it did not understand fully the results of applying `clean_names()` to the data.

We eventually sent it the following, explanining the data structure, after which it recovered:

```{markdown}
It wasn't working, so I did this (removed columns that broke the code)

# Clean and prepare
clean_df <- df |>
  clean_names() |>
  mutate(
    vaccine = factor(vaccine),
    geography_type = factor(geography_type),
    geography = factor(geography),
    season_survey_year = factor(season_survey_year),
    dimension_type = factor(dimension_type),
    dimension = factor(dimension)
  )

# Preview result
glimpse(clean_df)

and got back:

> glimpse(clean_df)
Rows: 220,729
Columns: 11
$ vaccine                <fct> "Seasonal Influenza", "Seasonal Influenza", "Seasonal Influenza", "Seasonal Influenza", "Seasonal Influenza", "Seasonal Influenza", "…
$ geography_type         <fct> Counties, Counties, Counties, Counties, Counties, Counties, States/Local Areas, States/Local Areas, States/Local Areas, States/Local …
$ geography              <fct> New Haven, New Haven, New Haven, New Haven, New Haven, New Haven, New Jersey, New Jersey, New Jersey, New Jersey, New Jersey, New Jer…
$ fips                   <chr> "09009", "09009", "09009", "09009", "09009", "09009", "34", "34", "34", "34", "34", "34", "34", "34", "34", "34", "34", "34", "34", "…
$ season_survey_year     <fct> 2018, 2021, 2020, 2021, 2018, 2019, 2009-10, 2009-10, 2009-10, 2009-10, 2009-10, 2009-10, 2009-10, 2009-10, 2009-10, 2009-10, 2009-10…
$ month                  <dbl> 1, 1, 1, 1, 1, 1, 5, 4, 11, 10, 9, 8, 9, 1, 12, 11, 10, 3, 2, 5, 4, 3, 2, 1, 12, 5, 5, 4, 3, 2, 12, 5, 8, 4, 3, 2, 1, 12, 11, 10, 9, …
$ dimension_type         <fct> >=18 Years, >=18 Years, Age, Age, Age, Age, Race and Ethnicity, Race and Ethnicity, Race and Ethnicity, Race and Ethnicity, Race and …
$ dimension              <fct> "Non-Medical Setting", "Non-Medical Setting", ">=18 Years", ">=18 Years", ">=18 Years", ">=18 Years", "American Indian or Alaska Nati…
$ estimate_percent       <chr> "45.5", "53.0", "52.4", "50.2", "34.0", "49.5", "57.7", "57.7", "34.7", "24.8", "10.8", "NR †", "NR †", "57.7", "54.8", "44.4", "NR †…
$ x95_percent_ci_percent <chr> "43.9 to 47.2", "46.0 to 60.9", "50.6 to 54.3", "45.4 to 55.8", "32.6 to 35.5", "47.8 to 51.3", "40.2 to 75.2 ‡", "40.2 to 75.2 ‡", "…
$ sample_size            <dbl> NA, NA, NA, NA, NA, NA, 48, 48, 552, 552, 552, 552, 48, 48, 48, 48, 48, 48, 48, 2554, 2554, 2554, 2554, 2554, 2554, 8306, 3107, 3107,…
> 

try again
```

This context helped if finds its way and we ended up with the following code block:


```{r}
#| message: false
#| warning: false

# Clean and prepare
clean_df <- df |>
  clean_names() |>
  mutate(
    vaccine = factor(vaccine),
    geography_type = factor(geography_type),
    geography = factor(geography),
    season_survey_year = factor(season_survey_year),
    dimension_type = factor(dimension_type),
    dimension = factor(dimension),
     estimate = as.numeric(str_replace(estimate_percent, "[^0-9.]", "")),
    ci_clean = str_remove(x95_percent_ci_percent, " ‡| †"),
    ci_lower = as.numeric(str_extract(ci_clean, "^[0-9.]+")),
    ci_upper = as.numeric(str_extract(ci_clean, "[0-9.]+$"))
  )

# Preview result
glimpse(clean_df)

```

### Data Summary

We then sent it this to inform it of the structure of the data, which is somewhat unusual.

```{markdown}
> table(clean_df$dimension)

                                             >=18 Years                                              >=6 Months 
                                                  26534                                                   10449 
                                             >=65 Years                                             13-17 Years 
                                                  10903                                                   10265 
                                            18-49 Years                                18-49 Years at High Risk 
                                                   9931                                                   11038 
                           18-49 Years not at High Risk                                             18-64 Years 
                                                   9257                                                    9942 
                               18-64 Years at High Risk                            18-64 Years not at High Risk 
                                                   9898                                                    8018 
                               25-64 Years at High Risk                 25-64 Years not in Initial Target Group 
                                                    496                                                     496 
                                             5-12 Years                                             50-64 Years 
                                                  10279                                                   11125 
                                    6 Months - 17 Years                                      6 Months - 4 Years 
                                                  12018                                                   10282 
6 Months - 64 Years at High Risk (Initial Target Group)          American Indian or Alaska Native, Non-Hispanic 
                                                    496                                                    1702 
                                    Asian, Non-Hispanic                                     Black, Non-Hispanic 
                                                   1712                                                   11067 
                                             Greater 65                               Greater than 18 Years flu 
                                                    770                                                     770 
                              Greater than 6 Months flu                                                Hispanic 
                                                    682                                                   11106 
                                        Medical Setting                                     Non-Medical Setting 
                                                    770                                                    6975 
                  Other or Multiple Races, Non-Hispanic                                          Pharmacy/Store 
                                                  11093                                                     770 
                                                 School                                     White, Non-Hispanic 
                                                    198                                                   11115 
                                              Workplace 
                                                    572 
> table(clean_df$dimension_type)

         >=18 Years          >=65 Years         18-49 Years         18-64 Years         50-64 Years 6 Months - 17 Years                 Age  Race and Ethnicity 
               6829                 416                 416                 416                 416                 792              163649               47795 

I tabulated the dimensions, and this is what you get.   Can you give me summarize code to make more sense of this that i can put in a notebook?

```

It returned us the following:

```{r summarize_dimensions}
dimension_type_summary <- clean_df %>% 
  count(dimension_type, name = "n") %>%                       # tally rows
  arrange(desc(n)) %>%                                        # largest first
  mutate(percent_of_total = n / sum(n))                       # share of entire data set

## 2 ── full breakdown: every dimension within each type -----------------------
dimension_breakdown <- clean_df %>% 
  count(dimension_type, dimension, name = "n") %>%            # tally rows
  group_by(dimension_type) %>% 
  mutate(
    percent_within_type = n / sum(n),                         # share inside the parent type
    rank_within_type   = dense_rank(desc(n))                  # rank for convenience
  ) %>% 
  arrange(dimension_type, desc(n)) %>% 
  ungroup()

## (optional) keep only the 5 largest dimensions per type ----------------------
top5_breakdown <- dimension_breakdown %>% 
  filter(rank_within_type <= 5) %>% 
  select(-rank_within_type)                                   # drop helper column

## ── Quick display ------------------------------------------------------------
dimension_type_summary %>% 
  kable(digits = 2, caption = "Records by dimension_type")

top5_breakdown %>% 
  kable(digits = 2, caption = "Top-5 dimensions within each dimension_type")
```

I then gave if this response:

```{markdown}
> # Display top 10 per type (optional, remove if you want all)
> dimension_summary |>
+   group_by(dimension_type) |>
+   slice_head(n = 10) |>
+   ungroup() |>
+   knitr::kable()

|dimension_type      |dimension                                      |     n|
|:-------------------|:----------------------------------------------|-----:|
|>=18 Years          |Non-Medical Setting                            |  6361|
|>=18 Years          |Medical Setting                                |   156|
|>=18 Years          |Pharmacy/Store                                 |   156|
|>=18 Years          |Workplace                                      |   156|
|>=65 Years          |Medical Setting                                |   104|
|>=65 Years          |Non-Medical Setting                            |   104|
|>=65 Years          |Pharmacy/Store                                 |   104|
|>=65 Years          |Workplace                                      |   104|
|18-49 Years         |Medical Setting                                |   104|
|18-49 Years         |Non-Medical Setting                            |   104|
|18-49 Years         |Pharmacy/Store                                 |   104|
|18-49 Years         |Workplace                                      |   104|
|18-64 Years         |Medical Setting                                |   104|
|18-64 Years         |Non-Medical Setting                            |   104|
|18-64 Years         |Pharmacy/Store                                 |   104|
|18-64 Years         |Workplace                                      |   104|
|50-64 Years         |Medical Setting                                |   104|
|50-64 Years         |Non-Medical Setting                            |   104|
|50-64 Years         |Pharmacy/Store                                 |   104|
|50-64 Years         |Workplace                                      |   104|
|6 Months - 17 Years |Medical Setting                                |   198|
|6 Months - 17 Years |Non-Medical Setting                            |   198|
|6 Months - 17 Years |Pharmacy/Store                                 |   198|
|6 Months - 17 Years |School                                         |   198|
|Age                 |>=18 Years                                     | 26534|
|Age                 |6 Months - 17 Years                            | 12018|
|Age                 |50-64 Years                                    | 11125|
|Age                 |18-49 Years at High Risk                       | 11038|
|Age                 |>=65 Years                                     | 10903|
|Age                 |>=6 Months                                     | 10449|
|Age                 |6 Months - 4 Years                             | 10282|
|Age                 |5-12 Years                                     | 10279|
|Age                 |13-17 Years                                    | 10265|
|Age                 |18-64 Years                                    |  9942|
|Race and Ethnicity  |White, Non-Hispanic                            | 11115|
|Race and Ethnicity  |Hispanic                                       | 11106|
|Race and Ethnicity  |Other or Multiple Races, Non-Hispanic          | 11093|
|Race and Ethnicity  |Black, Non-Hispanic                            | 11067|
|Race and Ethnicity  |Asian, Non-Hispanic                            |  1712|
|Race and Ethnicity  |American Indian or Alaska Native, Non-Hispanic |  1702|

Here is what I have:

Please notice that the age categories have overlap, so it's clear that these data are pre-aggregated, and we'll want to make sure we use mutually exclusive categories

Here is what I want:

I want to create a time series that shows the vacciation rate, by age, over time

Before you give me code, what do you reccomend
```

It went over challenges and recommendations.

We then asked it:

```
OK, give me that dataframe before we visualize
```

```{r}
mutually_exclusive_ages <- c(
  "6 Months - 4 Years",
  "5-12 Years",
  "13-17 Years",
  "18-49 Years",
  "50-64 Years",
  ">=65 Years"
)

# Build the cleaned time series dataframe
age_time_series <- clean_df |>
  filter(
    dimension_type == "Age",
    dimension %in% mutually_exclusive_ages,
    !is.na(estimate)
  ) |>
  group_by(season_survey_year, age_group = dimension) |>
  summarise(
    mean_estimate = mean(estimate, na.rm = TRUE),
    n_records = n(),
    .groups = "drop"
  ) |>
  mutate(
    age_group = fct_relevel(age_group, mutually_exclusive_ages)
  )

# View the resulting structure
glimpse(age_time_series)
```

I responded:

```{markdown}
> glimpse(age_time_series)
Rows: 85
Columns: 4
$ season_survey_year <fct> 2009-10, 2009-10, 2010-11, 2010-11, 2010-11, 2010-11, 2010-11, 2010-11, 2011-12, 2011-12, 2011-12, 2011-12, 2011-12, 2011-12, 2012-13, 20…
$ age_group          <fct> ">=65 Years", "50-64 Years", ">=65 Years", "13-17 Years", "18-49 Years", "5-12 Years", "50-64 Years", "6 Months - 4 Years", ">=65 Years",…
$ mean_estimate      <dbl> 46.28232, 38.60872, 52.04114, 30.16828, 25.02017, 43.74000, 35.90828, 51.23814, 51.03056, 28.14342, 23.20322, 43.89715, 33.89848, 54.0487…
$ n_records          <int> 1680, 1158, 598, 558, 575, 585, 580, 582, 602, 562, 590, 596, 592, 580, 614, 583, 606, 626, 607, 618, 603, 588, 592, 639, 590, 627, 616, …
> 

give me four lines for each group, annotated, by year
```

### Data Visualization

With the data now prepared and us asking for a line plot broken apart by age group, it gave us the following code, which worked the first time.

```{r}

ggplot(age_time_series, aes(x = season_survey_year, y = mean_estimate, group = age_group, color = age_group)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2) +
  labs(
    title = "Influenza Vaccination Coverage Over Time by Age Group",
    x = "Season/Survey Year",
    y = "Mean Vaccination Estimate (%)",
    color = "Age Group"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "bottom"
  )

```

This image needs refinement, which ChatGPT would happily provide if we continued, but it worked. We wrote zero code ourselves.