---
title: "AI & LLM Example"
author: "Erik Westlund"
date: "2025-06-11"
date-modified: "`r format(Sys.Date(), '%Y-%m-%d')`"
editor:
  render-on-save: true
---

## Overview

* We'll use an LLM to help us prepare data for a visualization
* Let's use the [CDC's data on influenza vaccination coverage for all ages](https://data.cdc.gov/Flu-Vaccinations/Influenza-Vaccination-Coverage-for-All-Ages-6-Mont/vh55-3he6/about_data)

## Data Preparation

```{r setup}
#| echo: false
#| message: false

# Install required packages if not already installed
required_packages <- c("dplyr", "ggplot2", "forcats", "janitor", "kableExtra", 
                      "readr", "stringr", "tidyr")
new_packages <- required_packages[!(required_packages %in% installed.packages()[,"Package"])]
if(length(new_packages)) install.packages(new_packages)

# Load required packages
library(dplyr)
library(ggplot2)
library(forcats)
library(janitor)
library(kableExtra)
library(readr)
library(stringr)
library(tidyr)

```

## Overview

* We'll use an LLM to help us prepare data for a visualization
* Let's use the [CDC's data on influenza vaccination coverage for all ages](https://data.cdc.gov/Flu-Vaccinations/Influenza-Vaccination-Coverage-for-All-Ages-6-Mont/vh55-3he6/about_data)

## Data Preparation

```{r}
#| message: false

df <- read_csv(here::here("data", "raw", "cdc_Influenza_Vaccination_Coverage_for_All_Ages__6__Months__20250610.csv"))

df |> glimpse()
```

## Chat with the LLM

### Data Preparation

From here, we can chat with the LLM to help us prepare the data for a visualization.

Below I provide key aspects of the prompts sent to the ChatGPT LLM as well as the results. I've excluded some of the LLM's missteps.

### Initial Data Exploration

I am working with an unfamiliar data set. Here is a glimpse using dplyr:

```{r}
df <- read_csv(here::here("data", "raw", "cdc_Influenza_Vaccination_Coverage_for_All_Ages__6__Months__20250610.csv"))
df |> glimpse()
```

The bot then summarized the data set.

Using the following prompt, we then sent the LLM the actual data file for more context (🚨 do not do this with private data! 🚨):

"I am actually going to give you the file. Interpret it. We are going to use dplyr and ggplot2 to prepare it and visualize aspects of it. Tell me what you see."

It responded with a summary, warning us about some of the odd structures of the data.

We then asked it to clean the data:

"Yes, let's get clean, standardized column names. Let's make all factor variables factor types. Let's extract the confidence intervals into separate columns. Let's just start there. Give me R code that I can paste into RStudio."

Please note from here on out, ChatGPT gave us code that includes library calls and new code. The library calls are loaded above.

It first gave us back broken code because it did not understand fully the results of applying `clean_names()` to the data.

We eventually sent it the following, explaining the data structure, after which it recovered:

```{r}
# Clean and prepare
clean_df <- df |>
  clean_names() |>
  mutate(
    vaccine = factor(vaccine),
    geography_type = factor(geography_type),
    geography = factor(geography),
    season_survey_year = factor(season_survey_year),
    dimension_type = factor(dimension_type),
    dimension = factor(dimension),
    estimate = as.numeric(str_replace(estimate_percent, "[^0-9.]", "")),
    ci_clean = str_remove(x95_percent_ci_percent, " ‡| †"),
    ci_lower = as.numeric(str_extract(ci_clean, "^[0-9.]+")),
    ci_upper = as.numeric(str_extract(ci_clean, "[0-9.]+$"))
  )

# Preview result
glimpse(clean_df)
```

### Data Summary

We then sent it this to inform it of the structure of the data, which is somewhat unusual:

```{r}
table(clean_df$dimension)
table(clean_df$dimension_type)
```

Here is what we have:

Please notice that the age categories have overlap, so it's clear that these data are pre-aggregated, and we'll want to make sure we use mutually exclusive categories.

Here is what we want:

We want to create a time series that shows the vaccination rate, by age, over time.

Before you give me code, what do you recommend?

It went over challenges and recommendations.

We then asked it:

"OK, give me that dataframe before we visualize"

```{r}
mutually_exclusive_ages <- c(
  "6 Months - 4 Years",
  "5-12 Years",
  "13-17 Years",
  "18-49 Years",
  "50-64 Years",
  ">=65 Years"
)

# Build the cleaned time series dataframe
age_time_series <- clean_df |>
  filter(
    dimension_type == "Age",
    dimension %in% mutually_exclusive_ages,
    !is.na(estimate)
  ) |>
  group_by(season_survey_year, age_group = dimension) |>
  summarise(
    mean_estimate = mean(estimate, na.rm = TRUE),
    n_records = n(),
    .groups = "drop"
  ) |>
  mutate(
    age_group = fct_relevel(age_group, mutually_exclusive_ages)
  )

# View the resulting structure
glimpse(age_time_series)
```

### Data Visualization

With the data now prepared and us asking for a line plot broken apart by age group, it gave us the following code, which worked the first time:

```{r}
ggplot(age_time_series, aes(x = season_survey_year, y = mean_estimate, group = age_group, color = age_group)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2) +
  labs(
    title = "Influenza Vaccination Coverage Over Time by Age Group",
    x = "Season/Survey Year",
    y = "Mean Vaccination Estimate (%)",
    color = "Age Group"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "bottom"
  )
```

This image needs refinement, which ChatGPT would happily provide if we continued, but it worked. We wrote zero code ourselves.

