---
title: "Application 3: Dot Plots"
author: "Erik Westlund"
date: "2025-06-12"

editor:
  render-on-save: true
---

```{r setup}
#| message: false

# List of required packages
required_packages <- c(
  "readr",
  "dplyr",
  "ggplot2",
  "forcats",
  "tidyr",
  "kableExtra",
  "stringr",
  "ggrepel",
  "maps",
  "usmap",
  "sf"
)

# Install missing packages
new_packages <- required_packages[!(required_packages %in% installed.packages()[,"Package"])]
if(length(new_packages)) install.packages(new_packages)

# Load all packages
for (package in required_packages) {
  library(package, character.only = TRUE)
}

source(here::here("examples", "colors.R"))
```

## Load and Prepare

As usual, we start by loading and preparing the data.

```{r data_prep}
#| message: false

data <- readr::read_csv(here::here("data", "processed", "simulated_data.csv"))
state_data <- readr::read_csv(here::here("data", "processed", "state_data.csv"))

data <- data |>
  mutate(
    state = as.factor(state),
    received_comprehensive_postnatal_care = as.numeric(received_comprehensive_postnatal_care),
    insurance = fct_relevel(insurance, "no_insurance"),
    race_ethnicity = fct_relevel(race_ethnicity, "white"),
    edu = fct_relevel(edu, "hs"),
    job_type = fct_relevel(job_type, "unemployed"),
  )

data$self_report_income <- factor(data$self_report_income, levels = c(
  "$0–$24,999",
  "$25,000–$49,999",
  "$50,000–$74,999",
  "$75,000–$99,999",
  "$100,000–$124,999",
  "$125,000–$149,999",
  "$150,000–$174,999",
  "$175,000+"
))

# Set "$50,000–$75,000" as the reference category
data$self_report_income <- fct_relevel(data$self_report_income, "$50,000–$75,000")

race_ethnicity_labels <- c("American Indian or Alaska Native" = "aian",
          "White" = "white",
          "Black" = "black",
          "Asian" = "asian",
          "Hispanic" = "hispanic",
          "Native Hawaiian or Pacific Islander" = "nhpi",
          "Other" = "other")

data |> head() |> kable()
```

## Dot Plots

In `applications_2_choropleth.qmd` we drew up a choropleth map of the receipt of comprehensive postnatal care by state. Let's see if a dot plot can help us understand the data better.

```{r dot-plot}
#| fig.height: 10

# Calculate overall proportions for all states
overall_care_data <- data |>
  group_by(state) |>
  summarize(
    proportion = mean(received_comprehensive_postnatal_care, na.rm = TRUE),
    n = n(),
    .groups = "drop"
  ) |>
  mutate(race_ethnicity = "Overall")

# Calculate proportions for each race
race_care_data <- data |>
  filter(race_ethnicity %in% c("white", "black", "hispanic", "asian")) |>
  group_by(state, race_ethnicity) |>
  summarize(
    proportion = mean(received_comprehensive_postnatal_care, na.rm = TRUE),
    n = n(),
    .groups = "drop"
  )  |>
  mutate(
    race_ethnicity = fct_recode(race_ethnicity, !!!race_ethnicity_labels)
  ) ## will have warning because we dropped some race/eth cats


# Combine overall and race-specific data
dot_plot_data <- bind_rows(overall_care_data, race_care_data)

# Sort states by overall proportion (to ensure consistent order across facets)
state_order <- dot_plot_data |>
  filter(race_ethnicity == "Overall") |>
  arrange(proportion) |>
  pull(state)

dot_plot_data <- dot_plot_data |>
  mutate(state = factor(state, levels = state_order)) |> 
  mutate(
    race_ethnicity = factor(
      race_ethnicity,
      levels = c("Overall", "White", "Black", "Hispanic", "Asian")
    )
  )

ggplot(dot_plot_data, aes(x = proportion, y = state)) +
  geom_point(size = 3, color = colors$blue$`500`) +
  geom_segment(
    aes(x = 0, xend = proportion, y = state, yend = state),
    color = colors$blue$`500`,
    linetype = "dotted"
  ) +
  facet_wrap(~race_ethnicity, ncol = 5, scales = "free_y") +
  scale_x_continuous(labels = scales::percent_format(accuracy = 1)) +
  labs(
    title = "Proportion Receiving Comprehensive Postnatal Care",
    x = "Proportion",
    y = "State",
    caption = "Source: Simulated Data"
  ) +
  theme_minimal() +
  theme(
    axis.text.y = element_text(size = 8),
    strip.text = element_text(size = 11, face = "bold"),
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
    plot.caption = element_text(size = 10, hjust = 0)
  )
```

We find that, in fact, the dot plot is superior to the choropleth map to visualize the spatial trends that exist in the data.

I would encourage you to always try to visualize spatial data using non-map-based visualizations. In many cases, maps foreground non-causal phenomena like population density.

For comparison, compare the above to the choropleth map we made in `applications_2_choropleth.qmd`, which we reproduce below:



```{r}
#| echo: false

us_map <- us_map(regions = "states")

rcp_by_state_data <- data |>
  group_by(state) |>
  summarize(
    proportion = mean(received_comprehensive_postnatal_care),
    n = n()
  ) |>
  ungroup()

# Merge summarized data with map data
rcp_by_state_data_mappable <- us_map |>
  left_join(
    rcp_by_state_data, 
    by = c("abbr"="state")
  )

ggplot(data = rcp_by_state_data_mappable) +
  geom_sf(aes(fill = proportion), color = "white") +
  scale_fill_continuous(
    low = colors$blue$`50`,
    high = colors$blue$`900`,
    name = "Proportion",
    labels = scales::percent,
  ) +
  labs(
    title = "Proportion Receiving Comprehensive Postnatal Care By State",
    caption = "Source: Simulated Data"
  ) +
  theme_minimal() +
  theme(
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    panel.grid = element_blank(),
    plot.title = element_text(hjust = 0.5, size = 12, face = "bold")
  )

# Will it work? Maps  by state
race_ethnicity_care_data <- data |>
  group_by(state, race_ethnicity) |>
  summarize(
    proportion = mean(received_comprehensive_postnatal_care, na.rm = TRUE),
    n = n(),
    	.groups = "drop"
  ) |>
  ungroup() |>
  mutate(
    race_ethnicity = fct_recode(race_ethnicity, !!!race_ethnicity_labels)
  )

race_ethnicity_care_sf <- us_map |>
  left_join(
	race_ethnicity_care_data,
	by = c("abbr" = "state")
  )

ggplot(race_ethnicity_care_sf) +
  geom_sf(aes(geometry = geom, fill = proportion), color = "white", size = 0.1) +
  scale_fill_continuous(
    low = colors$blue$`100`,
    high = colors$blue$`900`,
    name = "Proportion",
    labels = scales::percent,
    na.value = "grey90"
  ) +
  facet_wrap(~ race_ethnicity, ncol = 3, nrow = 3) +
  labs(
    title = "Proportion Receiving Comprehensive Postnatal Care by Race",
    fill = "Proportion",
    caption = "Source: Simulated Data"
  ) +
  theme_minimal() +
  theme(
    panel.grid = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    plot.title = element_text(hjust = 0.5, size = 12, face = "bold"),
    strip.text = element_text(size = 10)
  )

```