---
title: "PRAMS Data Analysis: Aggregation & Transformation"
author: "Erik Westlund"
date: "2025-06-11"
date-modified: "`r format(Sys.Date(), '%Y-%m-%d')`"

editor:
  render-on-save: true
---

```{r setup}

library(dplyr)
library(ggplot2)
library(forcats)
library(kableExtra)
library(readr)
library(ggtext)

source(here::here("examples", "colors.R"))
```

## Research Question

Let's examine the relationship between depression in the 3 months before birth and the number of previous live births, focusing specifically on Wisconsin.

## Data Preparation

First, let's load and prepare our data:

```{r load-data}

df_final <- readRDS(here::here("data", "processed", "cdc_prams_df_final.rds"))

# Filter for Wisconsin and relevant variables
df_wi <- df_final |>
  filter(
    location_abbr == "WI",
    subgroup_cat == "Number of Previous Live Births"
  ) |>
  select(
    location_abbr,
    subgroup,
    depression_within_3_months_birth
  )

# Let's look at the structure of our filtered data
df_wi |>
  glimpse()
```

Let's examine the unique values in our key variables:

```{r examine-variables}

# Check unique values in depression variable
df_wi |>
  select(depression_within_3_months_birth) |>
  distinct() |>
  kable()

# Check unique values in subgroup (number of previous live births)
df_wi |>
  select(subgroup) |>
  distinct() |>
  kable()
```

Now let's create our first visualization:

```{r first-plot}

p1 <- df_wi |>
  ggplot(aes(x = subgroup, y = depression_within_3_months_birth)) +
  geom_bar(
    stat = "identity", 
    fill = colors$blue[['100']],
    color = colors$blue[['400']],
    linewidth = 0.5,
    linetype = "solid"
  ) +
  geom_text(
    aes(label = sprintf("%.1f%%", depression_within_3_months_birth)),
    vjust = -1.2,
    size = 3.5
  ) +
  theme_minimal() +
  labs(
    title = "Depression Before Birth by Number of Previous Live Births",
    subtitle = "Wisconsin PRAMS Data",
    x = "Number of Previous Live Births",
    y = "Percentage Reporting Depression"
  ) +
  theme(
    plot.title = element_text(face = "bold"),
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank()
  )

p1
```

## Faceting By Small Multiples

A great strategy for showing patterns by groups is to facet by small multiples.

Let's start again with our data, not filtering by Wisconsin, but showing little bar charts for each state.

```{r facet-by-state}

p2 <- df_final |>
  filter(subgroup_cat == "Number of Previous Live Births") |>
  ggplot(aes(x = subgroup, y = depression_within_3_months_birth)) +
  geom_bar(
    stat = "identity", 
    fill = colors$blue[['100']],
    color = colors$blue[['400']],
    linewidth = 0.5,
    linetype = "solid"
  ) +
  geom_text(
    aes(label = sprintf("%.1f%%", depression_within_3_months_birth)),
    vjust = -1.2,
    size = 3.5
  ) +
  facet_wrap(~ location_abbr) +
  theme_minimal() +
  labs(
    title = "Depression Before Birth by Number of Previous Live Births",
    subtitle = "By State",
    x = "Number of Previous Live Births",
    y = "Percentage Reporting Depression"
  ) +
  theme(
    plot.title = element_text(face = "bold"),
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank(),
    strip.text = element_text(face = "bold")
  )

p2
```
This a start.  it has issues, though.  First, we have a lot of missing states. Let's fix that first.

```{r fix-missing-states}

p3 <- df_final |>
  filter(subgroup_cat == "Number of Previous Live Births") |>
  filter(!is.na(depression_within_3_months_birth)) |>
  ggplot(aes(x = subgroup, y = depression_within_3_months_birth)) +
  geom_bar(
    stat = "identity", 
    fill = colors$blue[['100']],
    color = colors$blue[['400']],
    linewidth = 0.5,
    linetype = "solid"
  ) +
  geom_text(
    aes(label = sprintf("%.1f%%", depression_within_3_months_birth)),
    vjust = -1.2,
    size = 3.5
  ) +
  facet_wrap(~ location_abbr) +
  theme_minimal() +
  labs(
    title = "Depression Before Birth by Number of Previous Live Births",
    subtitle = "By State",
    x = "Number of Previous Live Births",
    y = "Percentage Reporting Depression"
  ) +
  theme(
    plot.title = element_text(face = "bold"),
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank(),
    strip.text = element_text(face = "bold")
  )

p3


```

This is an improvement, but we see some issues.  First, our y-axis is too short, and annotations are getting cut off.  We can fix our y axis to go to 20

```{r fix-y-axis}

p4 <- p3 +
  scale_y_continuous(limits = c(0, 20))

p4
```

That's better.  Some stylistic adjustments:

* Need more space between the labels and the plot:


```{r style-adjustments}

p5 <- p4 +
  theme(
    axis.title.x = element_text(margin = margin(t = 10)),
    axis.title.y = element_text(margin = margin(r = 10))
  )

p5
```

Now, let's sort it by the *difference* between the birth. This requires data transformation, so we'll start from scratch.

```{r sort-by-difference}

# Calculate differences and sort states
state_diffs <- df_final |>
  filter(subgroup_cat == "Number of Previous Live Births") |>
  filter(!is.na(depression_within_3_months_birth)) |>
  group_by(location_abbr) |>
  summarize(
    first_birth = depression_within_3_months_birth[subgroup == "0"],
    later_births = mean(depression_within_3_months_birth[subgroup != "0"]),
    diff = first_birth - later_births
  ) |>
  arrange(desc(diff))

# Create ordered factor for states
state_order <- state_diffs$location_abbr

# Create plot with sorted states
p6 <- df_final |>
  filter(subgroup_cat == "Number of Previous Live Births") |>
  filter(!is.na(depression_within_3_months_birth)) |>
  mutate(location_abbr = factor(location_abbr, levels = state_order)) |>
  left_join(
    state_diffs |> select(location_abbr, diff),
    by = "location_abbr"
  ) |>
  ggplot(aes(x = subgroup, y = depression_within_3_months_birth)) +
  geom_bar(
    stat = "identity", 
    aes(fill = diff > 0, color = diff > 0),
    linewidth = 0.5,
    linetype = "solid"
  ) +
  scale_fill_manual(
    name = "Difference",
    values = c(
      "TRUE" = colors$blue[['100']],
      "FALSE" = colors$red[['100']]
    ),
    labels = c(
      "TRUE" = "Depression Improves With Extra Births",
      "FALSE" = "Depression Worsens With Extra Births"
    )
  ) +
  scale_color_manual(
    name = "Difference",
    values = c(
      "TRUE" = colors$blue[['400']],
      "FALSE" = colors$red[['400']]
    ),
    labels = c(
      "TRUE" = "Depression Improves With Extra Births",
      "FALSE" = "Depression Worsens With Extra Births"
    )
  ) +
  geom_text(
    aes(label = sprintf("%.1f%%", depression_within_3_months_birth)),
    vjust = -1.2,
    size = 3.5
  ) +
  facet_wrap(~ location_abbr) +
  scale_y_continuous(limits = c(0, 20)) +
  theme_minimal() +
  labs(
    title = "Depression Before Birth by Number of Previous Live Births",
    subtitle = "States ordered by difference between first birth and subsequent births",
    x = "Number of Previous Live Births",
    y = "Percentage Reporting Depression"
  ) +
  theme(
    plot.title = element_text(face = "bold"),
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank(),
    strip.text = element_text(face = "bold"),
    axis.title.x = element_text(margin = margin(t = 10)),
    axis.title.y = element_text(margin = margin(r = 10)),
    legend.position = "bottom",
    legend.title = element_blank()
  )

p6
```

This is interesting, but I wonder if we're using the right graph.

If we're interested in the *difference*, we can do a forest plot by state in a single plot. Let's start with a basic version.

```{r forest-plot}
#| message: false

# Calculate differences for forest plot
forest_data <- df_final |>
  filter(subgroup_cat == "Number of Previous Live Births") |>
  filter(!is.na(depression_within_3_months_birth)) |>
  group_by(location_abbr) |>
  summarize(
    first_birth = depression_within_3_months_birth[subgroup == "0"],
    later_births = mean(depression_within_3_months_birth[subgroup != "0"]),
    diff = first_birth - later_births
  ) |>
  arrange(diff)  # Sort by difference

# Create ordered factor for states
state_order <- forest_data$location_abbr

# Create forest plot
p7 <- forest_data |>
  mutate(location_abbr = factor(location_abbr, levels = state_order)) |>
  ggplot(aes(x = diff, y = location_abbr)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray50") +
  geom_segment(
    aes(
      x = 0,
      xend = diff,
      y = location_abbr,
      yend = location_abbr,
      color = diff > 0
    ),
    linewidth = 1
  ) +
  scale_color_manual(
    values = c(
      "TRUE" = colors$blue[['400']],
      "FALSE" = colors$red[['400']]
    ),
    labels = c(
      "TRUE" = "Depression Improves With Extra Births",
      "FALSE" = "Depression Worsens With Extra Births"
    )
  ) +
  scale_x_continuous(
    limits = c(min(forest_data$diff) - 2, max(forest_data$diff) + 2),
    breaks = seq(-10, 10, 2)
  ) +
  theme_minimal() +
  labs(
    title = "Difference in Depression Rates: First Birth vs. Subsequent Births",
    subtitle = "Positive values indicate higher depression rates for first births",
    x = "Difference in Depression Rate (First Birth - Subsequent Births)",
    y = NULL
  ) +
  theme(
    plot.title = element_text(face = "bold"),
    panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank(),
    legend.position = "bottom",
    legend.title = element_blank()
  )

p7
```

This is a decent start.  Let's add text annotations to the plot. Let's also use the `location` rather than the abbrevaition.

```{r add-text-annotations}
#| message: false

# Calculate differences for forest plot
forest_data <- df_final |>
  filter(subgroup_cat == "Number of Previous Live Births") |>
  filter(!is.na(depression_within_3_months_birth)) |>
  group_by(location_abbr, location) |>
  summarize(
    first_birth = depression_within_3_months_birth[subgroup == "0"],
    later_births = mean(depression_within_3_months_birth[subgroup != "0"]),
    diff = first_birth - later_births,
    .groups = "drop"
  ) |>
  arrange(diff)  # Sort by difference

# Create ordered factor for states
state_order <- forest_data$location

# Create forest plot
p8 <- forest_data |>
  mutate(location = factor(location, levels = state_order)) |>
  ggplot(aes(x = diff, y = location)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray50") +
  geom_segment(
    aes(
      x = 0,
      xend = diff,
      y = location,
      yend = location,
      color = diff > 0
    ),
    linewidth = 1
  ) +
  geom_text(
    aes(
      x = diff,
      label = sprintf("%+.1f%%", diff)
    ),
    hjust = ifelse(forest_data$diff > 0, -0.3, 1.3),
    size = 3.5,
    fontface = "bold"
  ) +
  scale_color_manual(
    values = c(
      "TRUE" = colors$blue[['400']],
      "FALSE" = colors$red[['400']]
    ),
    labels = c(
      "TRUE" = "Depression Improves With Extra Births",
      "FALSE" = "Depression Worsens With Extra Births"
    )
  ) +
  scale_x_continuous(
    limits = c(min(forest_data$diff) - 2, max(forest_data$diff) + 2),
    breaks = seq(-10, 10, 2)
  ) +
  theme_minimal() +
  labs(
    title = "Difference in Depression Rates: First Birth vs. Subsequent Births",
    subtitle = "Positive values indicate higher depression rates for first births",
    x = "Difference in Depression Rate (First Birth - Subsequent Births)",
    y = NULL
  ) +
  theme(
    plot.title = element_text(face = "bold"),
    panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank(),
    legend.position = "bottom",
    legend.title = element_blank()
  )

p8
```

The spacing is a little wide between the annotations. Let's also make them bold.

```{r forest-plot-style}
#| message: false

p9 <- forest_data |>
  mutate(location = factor(location, levels = state_order)) |>
  ggplot(aes(x = diff, y = location)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray50") +
  geom_segment(
    aes(
      x = 0,
      xend = diff,
      y = location,
      yend = location,
      color = diff > 0
    ),
    linewidth = 1
  ) +
  geom_text(
    aes(
      x = diff,
      label = sprintf("%+.1f%%", diff)
    ),
    hjust = ifelse(forest_data$diff > 0, -0.3, 1.3),
    size = 3.5,
    fontface = "bold"
  ) +
  scale_color_manual(
    values = c(
      "TRUE" = colors$blue[['400']],
      "FALSE" = colors$red[['400']]
    ),
    labels = c(
      "TRUE" = "Depression Improves With Extra Births",
      "FALSE" = "Depression Worsens With Extra Births"
    )
  ) +
  scale_x_continuous(
    limits = c(min(forest_data$diff) - 2, max(forest_data$diff) + 2),
    breaks = seq(-10, 10, 2)
  ) +
  theme_minimal() +
  labs(
    title = "Difference in Depression Rates: First Birth vs. Subsequent Births",
    subtitle = "Positive values indicate higher depression rates for first births",
    x = "Difference in Depression Rate (First Birth - Subsequent Births)",
    y = NULL
  ) +
  theme(
    plot.title = element_text(face = "bold"),
    panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank(),
    legend.position = "bottom",
    legend.title = element_blank()
  )

p9
```

Let's step back and think.

* The subtitle does the work of the caption. We can remove it and simplify the code.  
* The bold is too much.  Let's remove it.
* We need more top margin on our y axis.

```{r readjust-plot}
#| fig.height: 5

p10 <- forest_data |>
  mutate(location = factor(location, levels = state_order)) |>
  ggplot(aes(x = diff, y = location)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray50") +
  geom_segment(
    aes(
      x = 0,
      xend = diff,
      y = location,
      yend = location,
      color = ifelse(diff > 0, "positive", "negative")
    ),
    linewidth = 1
  ) +
  geom_text(
    aes(
      x = diff,
      label = sprintf("%+.1f%%", diff)
    ),
    hjust = ifelse(forest_data$diff > 0, -0.3, 1.3),
    size = 3.5
  ) +
  scale_color_manual(
    values = c(
      "positive" = colors$blue[['400']],
      "negative" = colors$red[['400']]
    )
  ) +
  scale_x_continuous(
    limits = c(min(forest_data$diff) - 2, max(forest_data$diff) + 2),
    breaks = seq(-10, 10, 2)
  ) +
  theme_minimal() +
  labs(
    title = "Difference in Depression Rates: First Birth vs. Subsequent Births",
    subtitle = "Positive values indicate higher depression rates for first births",
    x = "Difference in Depression Rate (First Birth - Subsequent Births)",
    y = NULL
  ) +
  theme(
    plot.title = element_text(face = "bold"),
    panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank(),
    legend.position = "none",
    axis.text.y = element_text(margin = margin(r = 10))
  )

p10
```
